---
title: "Class 10: Haloween Candy Mini Project"
author: "Barry"
format: gfm
---

## Background

Here we explore 538 Haloween candy data. They recently ran a rather large poll to determine which candy their readers like best. From their website: "While we don’t know who exactly voted, we do know this: 8,371 different IP addresses voted on about 269,000 randomly generated candy matchups".

Let's get the data. I will download to my project directory...

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## Winpercent
The most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy 

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
rownames(candy)
```

```{r}
candy["Milky Way",]$winpercent
```

```{r}
candy["Caramel Apple Pops",]
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

## A useful function from the skimr package

```{r}
skimr::skim(candy)
```

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10, col="red", fill="orange")
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.inds <- as.logical(candy$chocolate)
chocolate.wins <-  candy[chocolate.inds,]$winpercent
mean(chocolate.wins)
```

```{r}
fruit.inds <- as.logical(candy$fruity)
fruit.wins <- candy[fruit.inds,]$winpercent
mean(fruit.wins)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.wins, fruit.wins)
```

# 3. Candy Ranking


First setup some colors for different candy types


```{r}

my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "red"
#my_cols
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

```{r}
ggsave("tmp.png")
```

Now, for the first time, using this plot we can answer questions like:
> Q17. What is the worst ranked chocolate candy?
> Q18. What is the best ranked fruity candy?



## 4. Taking a look at pricepercent

What is the best (most liked in terms of `winpercent`) for the money (in terms of `pricepercent`)?

To answer this I will make a plot of winpercent vs pricepercent

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text()
  
```

The labels on this plot are too squished and overlapping. I am going to turn to the ggrepl package and the `geom_text_repel()` function to help avoid overlapping labels.

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

6. Principal Component Analysis

Let's do PCA on this dataset to get a low dimensional view that hopefully captures the essential essence of the data.

We will use the `prcomp()` function and set `scale=TRUE` because the `winpercent` and `pricepercent` values are on a different scale! 

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)

```


```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```


And a ggplot version

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])


p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```


And the loadings that live in the `$rotation` component of our PCA results

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2)
```



```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1, rownames(loadings)) +
  geom_col()
```




